---
title: "Homework 1: Loss Function"
author: "Di Gravio, Chiara"
date: "`r date()`"
output: htlm_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


#### Write functions that implement the L1 loss and tilted absolute loss functions.

```{r}
# L2 loss function
L2_loss <- function(y, yhat, tau){
  (y-yhat)^2}
# L1_loss function
L1_loss <- function(y, yhat, tau){
  abs(y-yhat)}
# L1 tilted
L1_tilted_loss <- function(y, yhat, tau){
  res <- c()
  for(i in 1:length(y)){
    if((y[i] - yhat[i]) > 0){res[i] <- tau*(y[i] - yhat[i])}
    if((y[i] - yhat[i]) <= 0){res[i] <- (tau - 1)*(y[i] - yhat[i])}
  }
  return(res)}
```

#### Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label (using the 'legend' function) the linear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.

```{r}
library(ElemStatLearn)

## load prostate data
data(prostate)
## subset to training examples
prostate_train <- subset(prostate, train=TRUE)

## fit simple linear model using numerical optimization (from prostate-data-lin.R)
fit_lin <- function(y, x, tau, loss=L2_loss, beta_init = c(-0.51, 0.75)) {
  # create the training error function
  err <- function(beta){
    # average loss on the training data
    mean(loss(y,  beta[1] + beta[2]*x, tau))
  }
  # optim by default minimises the fn argumenrt based on the par
  beta <- optim(par = beta_init, fn = err)
  return(beta)
}
## make predictions from linear model using the beta we fit to the data (from prostate-data-lin.R)
predict_lin <- function(x, beta){
  beta[1] + beta[2]*x
}
  

# compute parameters that minimises the loss function
lin_beta_L2 <- fit_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L2_loss, tau = NULL)
lin_beta_L1 <- fit_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_loss, tau = NULL)
lin_beta_tiltedL1_025 <- fit_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_tilted_loss, tau = 0.25)
lin_beta_tiltedL1_075 <- fit_lin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_tilted_loss, tau = 0.75)

# compute predicted values for each loss function
x_grid <- seq(min(prostate_train$lpsa), max(prostate_train$lpsa), length.out=100)
lin_pred_L2 <- predict_lin(x = x_grid, beta = lin_beta_L2$par)
lin_pred_L1 <- predict_lin(x = x_grid, beta = lin_beta_L1$par)
lin_pred_tiltedL1_025 <- predict_lin(x = x_grid, beta = lin_beta_tiltedL1_025$par)
lin_pred_tiltedL1_075 <- predict_lin(x = x_grid, beta = lin_beta_tiltedL1_075$par)

# plot the values from the dataset
plot(prostate_train$lpsa, prostate_train$lcavol, xlab="log Prostate Screening Antigen (psa)", 
     ylab="log Cancer Volume (lcavol)", ylim = c(-2, 5))
# add lines for the predicted values
lines(x = x_grid, y = lin_pred_L2, col = "black")
lines(x = x_grid, y = lin_pred_L1, col = "red")
lines(x = x_grid, y = lin_pred_tiltedL1_025, col = "blue")
lines(x = x_grid, y = lin_pred_tiltedL1_075, col = "green")

# add legend to the plot
legend("bottomright", legend=c("L2 loss", "L1 loss", "Tilted L1 (tau = 0.25)", "Tilted L1 (tau = 0.75)"),
       col=c("black", "red", "blue", "green"), lty = 1, cex=0.8)
```

#### Write functions to fit and predict from a simple exponential (nonlinear) model with three parameters defined by $beta[1] + beta[2] \times exp(-beta[3] \times x)$. Hint: make copies of 'fit_lin' and 'predict_lin' and modify them to fit the nonlinear model. Use c(-1.0, 0.0, -0.3) as 'beta_init'.

```{r}
## fit simple nonlinear model using numerical optimization
fit_nonlin <- function(y, x, tau, loss, beta_init = c(-1.0, 0.0, -0.3)) {
  # create the training error function
  err <- function(beta){
    # average loss on the training data
    mean(loss(y,  beta[1] + beta[2]*exp(-beta[3]*x), tau))
  }
  # optim by default minimises the fn argumenrt based on the par
  beta <- optim(par = beta_init, fn = err, control=list(maxit= 1e4))
  return(beta)
}
## make predictions from linear model using the beta we fit to the data
predict_nonlin <- function(x, beta){
  beta[1] + beta[2]*exp(-beta[3]*x)
}
  
```

#### Create a figure that shows lpsa (x-axis) versus lcavol (y-axis). Add and label (using the 'legend' function) the nonlinear model predictors associated with L2 loss, L1 loss, and tilted absolute value loss for tau = 0.25 and 0.75.

```{r}
# compute parameters that minimises the loss function
nonlin_beta_L2 <- fit_nonlin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L2_loss, tau = NULL)
nonlin_beta_L1 <- fit_nonlin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_loss, tau = NULL)
nonlin_beta_tiltedL1_025 <- fit_nonlin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_tilted_loss, tau = 0.25)
nonlin_beta_tiltedL1_075 <- fit_nonlin(y=prostate_train$lcavol, x=prostate_train$lpsa, loss=L1_tilted_loss, tau = 0.75)

# compute predicted values for each loss function
x_grid <- seq(min(prostate_train$lpsa), max(prostate_train$lpsa), length.out=100)
nonlin_pred_L2 <- predict_nonlin(x = x_grid, beta = nonlin_beta_L2$par)
nonlin_pred_L1 <- predict_nonlin(x = x_grid, beta = nonlin_beta_L1$par)
nonlin_pred_tiltedL1_025 <- predict_nonlin(x = x_grid, beta = nonlin_beta_tiltedL1_025$par)
nonlin_pred_tiltedL1_075 <- predict_nonlin(x = x_grid, beta = nonlin_beta_tiltedL1_075$par)

# plot the values from the dataset
plot(prostate_train$lpsa, prostate_train$lcavol, xlab="log Prostate Screening Antigen (psa)", 
     ylab="log Cancer Volume (lcavol)", ylim = c(-2, 5))
# add lines for the predicted values
lines(x = x_grid, y = nonlin_pred_L2, col = "black")
lines(x = x_grid, y = nonlin_pred_L1, col = "red")
lines(x = x_grid, y = nonlin_pred_tiltedL1_025, col = "blue")
lines(x = x_grid, y = nonlin_pred_tiltedL1_075, col = "green")

# add legend to the plot
legend("bottomright", legend=c("L2 loss", "L1 loss", "Tilted L1 (tau = 0.25)", "Tilted L1 (tau = 0.75)"),
       col=c("black", "red", "blue", "green"), lty = 1, cex=0.8)
```





